Resonant Scattering in a Uniform Sphere with Large Optical Depth

### 2021-10-05 ms.tex
Incorporating Shane's edits to introduction

The outer layers of planetary atmospheres are central to a planet's evolution, since they can shelter the lower atmosphere from high energy radiation as well as regulate the escape of gas into space. Ionization and subsequent recombination can generate \lya within the planet's atmosphere, dissociating molecules in lower layers and creating pressure forces that drive an outflow of escaping H atoms. \lya can also excite H atoms to the 2p state, creating a population of Balmer-line absorbers that can be observed via transmission spectroscopy \citep{2017ApJ...851..150H}.

@@

Numerical simulations can be used to understand the complex layers of these irradiated exoplanet atmospheres. However, the extreme optical depths at \lya line center introduce a computational barrier to simulating radiative transfer via Monte Carlo methods. The main impediment to performing magnetohydrodynamic simulations with fully a coupled radiative transfer solution via Monte Carlo methods is the prohibitive computational cost. The number of scatterings a photon undergoes is proportional to the line center optical depth, $\tau_0$.  Near the base of the atmosphere, the line center optical depth is ${\sim}10^9$, so most of the time is spent following photons in these cells. A method that can treat these zones via a simple solution rather than Monte Carlo has the potential to greatly accelerate the calculation. A complete, general analytic solution for the resonance scattering has never been derived but approximate solutions exist in certain limits.

\citet{1973MNRAS.162...43H} demonstrated in slab geometry that resonance-line radiation transfer in optically-thick media can be described by the Poisson equation. However, their solution contains an incorrect treatment of the boundary condition. To our knowledge, the errors introduced by this treatment have never been quantified. They attempt a separation of variables $J(\tau,\sigma) = \theta(\tau) j(\sigma)$ in spatial variable $\tau$ and frequency variable $\sigma$ (their Equations 16 and 23). The solutions for the eigenfunctions $\theta(\tau)$ and $j(\sigma)$ then depend explicitly on the separation constant $\lambda$. In order to satisfy the boundary conditions, the separation constant must satisfy an eigenvalue equation of the form
\be
\lambda \tan(\lambda B) & = & \frac{3}{2} \phi \Delta,
\label{eq:evalue}
\ee
where $2B$ is the slab optical depth at line center, and $\phi$ is the line profile. The key point is that the line profile depends on one of the coordinates, frequency, and this causes the eigenvalues of the separation constant to depend on frequency. Thus, the separation constant is not constant, and it does not satisfy the Poisson equation since the frequency derivatives will act on the separation ``constant", giving extra terms. In the limit of large optical depth $B$, they approximate the eigenvalues as $\lambda_n B \simeq \pi (n-1/2)$. However, this solution gives zero mean intensity at the surface. Their Eq. 34 subsequently allowed the separation constant to have a small deviation from the above expression, which was explicitly frequency dependent. This allowed a nonzero intensity at the surface, but at the cost of rendering the separation of variables assumption invalid. Our treatment, using the correct boundary condition, quantifies the errors in their ansatz.

Several other works have followed \citet{1973MNRAS.162...43H}, building upon their result without addressing the incorrect boundary condition. \citet{1990ApJ...350..216N} attempts to extend the solution to media of intermediate optical depth, including the effects of scattering in the Doppler core of the line, but again does not use the correct boundary condition. \citet{2006ApJ...649...14D} generalize the same problem to spherical geometry, as is used here. Following this, \citet{2015MNRAS.449.4336S} employ a ``core skipping'' method to handle photon frequency redistribution within the Doppler core. This technique involves drawing the atom's velocity from a truncated Gaussian distribution, forcing photons back into the wing when they scatter into the core. % This is motivated by the assumption that all core scattering results in zero spatial diffusion until the photon is redistributed back out into the wing where the mean free path becomes much larger and it can diffuse in space again. However, it is not clear whether this prescription is fully independent of the choice of velocity distribution.

>>>>

The outer layers of planetary atmospheres are central to a planet's evolution, since they can shelter the lower atmosphere from high energy radiation as well as regulate the escape of gas into space. Ionization and subsequent recombination can generate \lya within the planet's atmosphere, dissociating molecules in lower layers and creating pressure forces that drive an outflow. \lya can also excite H atoms to the 2p state, creating a population of Balmer-line absorbers that can be observed via transmission spectroscopy \citep{2017ApJ...851..150H}.

@@

Numerical simulations can be used to understand the dynamics of these irradiated exoplanet atmospheres. The extreme optical depths at \lya line center impose a large computational cost for solving radiative transfer with Monte Carlo methods directly coupled with magnetohydrodynamic simulations. The number of scatterings a photon undergoes is proportional to the line center optical depth, $\tau_0$.  Near the base of the atmosphere, the line center optical depth is ${\sim}10^9$, so most of the time is spent following photons in these cells. A method that can treat these zones incorporating a direct solution rather than traditional Monte Carlo has the potential to greatly accelerate the calculation.

A complete, general analytic solution for the resonance scattering has never been derived but approximate solutions exist in certain limits. \citet{1973MNRAS.162...43H} demonstrated in slab geometry that resonance-line radiation transfer in optically-thick media can be described by the Poisson equation. However, their solution contains an incorrect treatment of the boundary condition. To our knowledge, the errors introduced by this treatment have never been quantified. They attempt a separation of variables $J(\tau,\sigma) = \theta(\tau) j(\sigma)$ in spatial variable $\tau$ and frequency variable $\sigma$ (their Equations 16 and 23). The solutions for the eigenfunctions $\theta(\tau)$ and $j(\sigma)$ then depend explicitly on the separation constant $\lambda$. In order to satisfy the boundary conditions, the separation constant must satisfy an eigenvalue equation of the form
\be
\lambda \tan(\lambda B) & = & \frac{3}{2} \phi \Delta,
\label{eq:evalue}
\ee
where $2B$ is the slab optical depth at line center, $\phi$ is the line profile, and $\Delta$ is the Doppler width. The key point is that the line profile depends on one of the coordinates, frequency, and this causes the eigenvalues of the separation constant to depend on frequency. Thus, the separation constant is not constant, and the solution does not satisfy the Poisson equation since the frequency derivatives will act on the separation ``constant", giving extra terms. In the limit of large optical depth $B$, they approximate the eigenvalues as $\lambda_n B \simeq \pi (n-1/2)$. However, this solution gives zero mean intensity at the surface. Their Eq. 34 subsequently allowed the separation constant to have a small deviation from the above expression, which was explicitly frequency dependent. This allowed a nonzero intensity at the surface, but at the cost of rendering the separation of variables assumption invalid. Our treatment, using the correct boundary condition, quantifies the errors in their ansatz.

Several other works have followed \citet{1973MNRAS.162...43H}, building upon their result without addressing the incorrect boundary condition. \citet{1990ApJ...350..216N} attempts to extend the solution to media of intermediate optical depth, including the effects of scattering in the Doppler core of the line, but again does not use the correct boundary condition. \citet{2006ApJ...649...14D} generalize the same problem to spherical geometry, as is used here.
